Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-12 ind_files/01_Introduction.tex
Move math background into separate chapter for clarity

Chapter \ref{chap_BD} deals with the evolution of discrete traits. In this case, the system is finite-dimensional, since we can completely specify the state of the system by simply listing out the number of individuals of each type in a vector. To facilitate readership by a broad audience, I only assume passing familiarity with calculus (derivatives, integrals, Taylor expansions) and probability. Familiarity with stochastic calculus is helpful for some sections but is not required. I present a brief introduction to the relevant mathematics in section \ref{sec_math_background} and present a toy example of tracking population size of a population of identical individuals in section \ref{sec_1D_processes}. I introduce a description of the system via a `master equation', and then conduct a `system-size expansion' to obtain a Fokker-Planck equation for the system. Finally, I conduct a weak noise approximation to arrive at a linear Fokker-Planck equation which can be solved exactly using some stochastic calculus to arrive at a closed-form solution given by a time-dependent Ornstein-Uhlenbeck process, thus illustrating all the major tools required. In section \ref{sec_nD_processes}, I present a multivariate process to describe the evolution of discretely varying traits, and, as before, use the system size expansion to arrive at a continuous description of change in trait frequencies as an SDE under mild assumptions on the functional forms of the birth and death rates. Unlike many classic stochastic formulations in evolutionary theory~\citep{fisher_genetical_1930,wright_evolution_1931,moran_random_1958,crow_introduction_1970, lande_natural_1976}, I do not assume a fixed (effective) population size and instead allow the total population size to fluctuate over time. I show that the deterministic limit of this process is the well-known replicator-mutator equation (or equivalently, the Price equation), thus establishing the microscopic basis of well-known equations from stochastic first principles. I also illustrate some general predictions that can be made using the weak noise approximation for the sake of completeness. While the mathematics of section \ref{sec_nD_processes} is standard and well-understood, it has, to the best of my knowledge, not been used in this context in the generality we use here. Several specific models of specific systems do use these mathematical techniques, but these papers are often written assuming familiarity with notions in physics and/or mathematics and thus may not be very accessible to theoretical ecologists who do not have formal training in these subjects (but see~\cite{czuppon_understanding_2021} for a recent pedagogical review on the general approach as applied to Wright-Fisher and Moran processes, where total population size is constant). As such, chapter \ref{chap_BD} also serves as a tutorial and technical introduction to the major ideas - For ecologists, the chapter introduces `system size expansions' and illustrates their use in a general setting, and can be seen as a tutorial on modelling finite populations analytically with minimal assumptions; For population geneticists, the chapter illustrates how system-size approximations (`diffusion approximations' in the population genetics literature) can be carried out without assuming a constant (effective) population size and how this generalization has important consequences for the evolutionary forces at play; For physicists and applied mathematicians, the chapter presents a study of the consequences of applying the system-size expansion to the kind of density-dependent birth-death processes that are widely applicable in ecology and evolution.

@@

Part \ref{part_summary} summarizes the major results of our formalism and presents some simple equations that can be argued to be `fundamental theorems' of population biology in the sense of~\cite{queller_fundamental_2017} or `unifying factors' in the sense of~\cite{lion_theoretical_2018}. These equations reduce to well-known results such as the Price equation, the replicator-mutator equation from evolutionary game theory, and Fisher's fundamental theorem from population genetics in the infinite population limit. For finite populations, these same equations predict a new evolutionary force, `noise-induced selection', that has still not found its way into the standard formal canon of evolutionary biology and whose significance is only recently being recognized~\citep{constable_demographic_2016,mcleod_social_2019,mazzolini_universality_2022, kuosmanen_turnover_2022}. Implications of noise-induced selection are also discussed in part \ref{part_summary}. Readers who are averse to or do not care for mathematical details can safely skip part \ref{part_theory} entirely and directly read part \ref{part_summary} for the major takeaways of this thesis, though I strongly recommend working through the formalism properly if possible.

>>>>

Chapter \ref{chap_math_background} provides the basic mathematical background required and illustrates the major ideas that we will use. To facilitate readership by a broad audience, I only assume passing familiarity with calculus (derivatives, integrals, Taylor expansions) and probability. Familiarity with stochastic calculus is helpful for some sections but is not required. I present a brief introduction to the relevant mathematics in section \ref{sec_math_background} and present a toy example of tracking population size of a population of identical individuals in section \ref{sec_1D_processes}. I introduce a description of the system via a `master equation', and then conduct a `system-size expansion' to obtain a Fokker-Planck equation for the system. Finally, I conduct a weak noise approximation to arrive at a linear Fokker-Planck equation which can be solved exactly using some stochastic calculus to arrive at a closed-form solution given by a time-dependent Ornstein-Uhlenbeck process, thus illustrating all the major tools required.

Chapter \ref{chap_BD} deals with the evolution of discrete traits. In this case, the system is finite-dimensional, since we can completely specify the state of the system by simply listing out the number of individuals of each type in a vector. I introduce a general multivariate process to describe the evolution of discretely varying traits, and use the system size expansion to arrive at a continuous description of change in trait frequencies as an SDE under mild assumptions on the functional forms of the birth and death rates. Unlike many classic stochastic formulations in evolutionary theory~\citep{fisher_genetical_1930,wright_evolution_1931,moran_random_1958,crow_introduction_1970, lande_natural_1976}, I do not assume a fixed (effective) population size and instead allow the total population size to fluctuate over time. I show that the deterministic limit of this process is the well-known replicator-mutator equation (or equivalently, the dynamic version of the Price equation), thus establishing the microscopic basis of well-known equations from stochastic first principles. I also illustrate some general predictions that can be made using the weak noise approximation for the sake of completeness. While the mathematics of this chapter is standard and well-understood, it has, to the best of my knowledge, not been used in this context in the generality we use here. Several specific models of specific systems do use these mathematical techniques, but these papers are often written assuming familiarity with notions in physics and/or mathematics and thus may not be very accessible to theoretical ecologists who do not have formal training in these subjects (but see~\cite{czuppon_understanding_2021} for a recent pedagogical review on the general approach as applied to Wright-Fisher and Moran processes, where total population size is constant). As such, chapters \ref{chap_math_background} and \ref{chap_BD} together also serve as a tutorial and technical introduction to some theoretical ideas: For ecologists, the chapter introduces `system size expansions' and illustrates their use in a general setting, and can be seen as a tutorial on modelling finite populations analytically with minimal assumptions; For population geneticists, the chapter illustrates how system-size approximations (`diffusion approximations' in the population genetics literature) can be carried out without assuming a constant (effective) population size and how this generalization has important consequences for the evolutionary forces at play; For physicists and applied mathematicians, the chapter presents a study of the consequences of applying the system-size expansion to the kind of density-dependent birth-death processes that are widely applicable in ecology and evolution - Unlike many physical systems, though processes occur in terms of numbers or densities, predictions in evolution are typically in terms of frequencies of types, and this fact has subtle consequences that can be overlooked if one only works with densities.

@@

Part \ref{part_summary} summarizes the major results of our formalism and presents some simple equations that can be argued to be `fundamental theorems' of population biology in the sense of~\cite{queller_fundamental_2017}, which together form a `unifying perspective' in the sense of~\cite{lion_theoretical_2018}. These equations reduce to well-known results such as the Price equation, the replicator-mutator equation from evolutionary game theory, and Fisher's fundamental theorem from population genetics in the infinite population limit. For finite populations, these same equations predict a new evolutionary force, `noise-induced selection', that has still not found its way into the standard formal canon of evolutionary biology and whose significance is only recently being recognized~\citep{constable_demographic_2016,mcleod_social_2019,mazzolini_universality_2022, kuosmanen_turnover_2022}. Implications of noise-induced selection are also discussed in part \ref{part_summary}. Readers who are averse to or do not care for mathematical details can safely skip part \ref{part_theory} entirely and directly read part \ref{part_summary} for the major takeaways of this thesis, though I strongly recommend working through the formalism properly if possible.
